A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-07-11 preliminaries.tex
Grammar check
some clean-up

A monoid $\Monoid = \Tuple{X;\MonoidOp;0_{\Monoid}}$ is an algebraic structure
consisting of the non-empty carrier set of elements, $X$, an associative binary operation
$X \times X \rightarrow X$ denoted as $\MonoidOp$, that is where for all $a, b,
c \in X$, $(a \MonoidOp b) \MonoidOp c = a \MonoidOp (b \MonoidOp c)$. Finally,
$\Monoid$ must have an identity element $0_{\Monoid} \in X$ such that
$0_{\Monoid} \MonoidOp a = a \MonoidOp 0_{\Monoid} =   a$ for every $a \in X$.
We sometimes use integer multiplication to represent repeated application of
$\MonoidOp$, e.g. $3a = a \MonoidOp a \MonoidOp a$, for $a \in X$. $\Monoid$ is
called \textit{commutative} if $\MonoidOp$ also commutes, that is if $a
\MonoidOp b = b \MonoidOp a$ for all $a, b \in X$.

Finally, a \textit{homomorphism} is a structure-preserving map between two monoids
$\Monoid_1, \Monoid_2$, that is a map $\Map : S_1 \rightarrow S_2$ such that $\Map(a
\MonoidOp_1 b) = \Map(a) \MonoidOp_2 \Map(b)$, where $S_1, S_2$ are the carrier
sets of $\Monoid_1$ and $\Monoid_2$ respectively, and $\MonoidOp_1, \MonoidOp_2$
their binary operations.

@@

We define an alphabet as a finite set of symbols $\Alphabet$ with words $\Strings$, and
the concatenation operation as $s_1 \Concat{} s_2$ over two strings $s_1, s_2$.
Note that $\Strings = \Tuple{\Alphabet;\Concat{};\epsilon}$, is a
non-commutative monoid, referred to as the free monoid. The string length
function, $\Length{s}$ is an example of a homomorphism between $\Strings$
and~$\mathbb{Z}$.

An automaton~$\Automaton$ with alphabet~$\Alphabet$ is
$\AutomatonTuple$ where $\Transitions \subseteq \States \times \Alphabet \times
\States$, $\States$ is its states, $\InitialState$ its
initial state, and $\AcceptingStates$ is its set of accepting states.  We
write a transition $\Transition = \Tuple{\State, \Label, \State'} \in
\Transitions$ as $\Transition = \FromLabelTo{\State}{\Label}{\State'}$.
Similarly, we use the notation $\FromLabelTo{\State}{}{}$ to refer to the set of
transitions starting in $\State$, and $\FromLabelTo{}{}{\State}$ to refer to the
set of transitions coming into $\State$, whenever the automaton is clear from
the context.

We will let variables $\Transition, \Transition', \Transition_1, \ldots,
\Transition_n$ etc describe transitions, $\State, \ldots, \State_n$ states, and
$\Automaton, \ldots, \Automaton_n$ automata, and use subscript indexing
($\Transitions_\Automaton$) to refer to the transitions, states, etc of a given
automaton.

By a \emph{product} of two automata $\Automaton_1, \Automaton_2$, written
$\Automaton_1 \times \Automaton_2$, we mean an automaton constructed to run
$\Automaton_1$ and $\Automaton_2$ in parallel on an input and only accept the
input if both automata would do so.

We refer to the resulting product states as tuples, $\Tuple{\State, \State'}$,
which represent the state of the product automaton where $\Automaton_1$ would be
in $\State$ and $\Automaton_2$ would be in $\State'$. Note that since we use
ordered tuples the product is technically (but w.l.o.g) not commutative; the
left-hand-side must come from the left-hand term. The sole purpose of this
matching is to allow us to speak with precision about the origins of components
in a product.

@@

% AS: Reviewers suggest we drop this:
% PR, shortened a bit
Parikh's theorem states that any context-free language has a Parikh-equivalent
regular language (c.f.~\cite{construction} for a construction of such automata
from context-free grammars and~\cite{bounds} for bounds on its size).
%However,
%there are languages that are not context-free that also have semilinear images
%under~$\ParikhMap$ (e.g. $\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) =
%\ParikhMap((abc)^*) = \CountOf{a} = \CountOf{b} = \CountOf{c} \land \CountOf{a}
%\geq 0$). This means they can be represented as a quantifier-free Presburger
%formula.
%
While Parikh's theorem applies to arbitrary context-free languages, in this
paper we focus only on regular languages.

>>>>

A monoid $\Monoid = \Tuple{X;\MonoidOp;0_{\Monoid}}$ is an algebraic structure
consisting of the non-empty carrier set of elements, $X$; an associative binary operation
$X \times X \rightarrow X$ denoted as $\MonoidOp$, that is where for all $a, b,
c \in X$, $(a \MonoidOp b) \MonoidOp c = a \MonoidOp (b \MonoidOp c)$; 
and an identity element $0_{\Monoid} \in X$ such that
$0_{\Monoid} \MonoidOp a = a \MonoidOp 0_{\Monoid} =   a$ for every $a \in X$.
We sometimes use integer multiplication to represent a repeated application of
$\MonoidOp$, e.g. $3a = a \MonoidOp a \MonoidOp a$, for $a \in X$. $\Monoid$ is
called \textit{commutative} if $\MonoidOp$ also commutes, that is if $a
\MonoidOp b = b \MonoidOp a$ for all $a, b \in X$.

A \textit{homomorphism} is a structure-preserving map between two
monoids $\Monoid = \Tuple{X;\MonoidOp;0_{\Monoid}}$ and
$\Monoid' = \Tuple{X';\MonoidOp';0_{\Monoid'}}$, that is a map
$\Map : X_1 \rightarrow X_2$ such that
$\Map(a \MonoidOp b) = \Map(a) \MonoidOp' \Map(b)$ and
$\Map(0_{\Monoid}) = 0_{\Monoid'}$.

@@

We define an alphabet as a finite set of symbols $\Alphabet$ with words $\Strings$, and
the concatenation operation as $s_1 \Concat{} s_2$ over two strings $s_1, s_2$.
Note that $\Strings = \Tuple{\Alphabet;\Concat{};\epsilon}$, is a
non-commutative monoid, referred to as the free monoid on $\Alphabet$.
The string length
function, $\Length{s}$ is an example of a homomorphism between $\Strings$
and~$\mathbb{Z}$.

A \emph{finite-state automaton}~$\Automaton$ with alphabet~$\Alphabet$ is a
tuple~$\AutomatonTuple$, where $\States$ is the set of states,
$\InitialState$ the initial state, $\AcceptingStates$ the set of
accepting states, and
$\Transitions \subseteq \States \times \Alphabet \times \States$ the
transition relation.  We write a transition
$\Transition = \Tuple{\State, \Label, \State'} \in \Transitions$ as
$\Transition = \FromLabelTo{\State}{\Label}{\State'}$.  Similarly, we
use the notation $\FromLabelTo{\State}{}{}$ to refer to the set of
transitions starting in $\State$, and $\FromLabelTo{}{}{\State}$ to
refer to the set of transitions coming into $\State$, whenever the
automaton is clear from the context. An accepting run of $\Automaton$
is a sequence
$\InitialState = q_0 \xrightarrow{l_1} q_1 \xrightarrow{l_2} \cdots
\xrightarrow{l_k} q_k \in \AcceptingStates$, and the accepted word is
$l_1 l_2 \ldots l_k \in \Strings$. The set of words accepted by an
automaton~$\Automaton$ is denoted by
$\Language(\Automaton) \subseteq \Strings$.

We will let variables $\Transition, \Transition', \Transition_1, \ldots,
\Transition_n$, etc., denote transitions, $\State, \ldots, \State_n$ states, and
$\Automaton, \ldots, \Automaton_n$ automata, and use subscript indexing
($\Transitions_\Automaton$) to refer to the transitions, states, etc., of a given
automaton.

The \emph{product} of two
automata~$\Automaton_1 = \Tuple{\States^1, \InitialState^1,
  \AcceptingStates^1, \Transitions^1}$ and
$\Automaton_2 = \Tuple{\States^2, \InitialState^2,
  \AcceptingStates^2, \Transitions^2}$ is the automaton
\begin{align*}
\Automaton_1 \times \Automaton_2 &=
\Tuple{\States_1 \times \States_2, \STuple{\InitialState^1, \InitialState^2},
\AcceptingStates^1 \times \AcceptingStates^2,
  \Transitions}\\
  \text{with}\quad \Transitions &= \{\STuple{\STuple{q,q''}, l, \STuple{q',q'''}} \mid \STuple{q, l, q'} \in \Transitions^1, \STuple{q'', l, q'''} \in \Transitions^2\}~.
\end{align*}
The product automaton runs $\Automaton_1$ and $\Automaton_2$ in
parallel on an input and only accepts the input if both automata would
do so; we have $\Language(\Automaton_1 \times \Automaton_2) =
\Language(\Automaton_1) \cap \Language(\Automaton_2)$.

@@

% AS: Reviewers suggest we drop this:
% PR, shortened a bit
Parikh's theorem states that any context-free language has a Parikh-equivalent
regular language (c.f.~\cite{construction} for a construction of such automata
from context-free grammars and~\cite{bounds} for bounds on its size).
The Parikh image is therefore a semi-linear set and Presburger-definable.
%However,
%there are languages that are not context-free that also have semilinear images
%under~$\ParikhMap$ (e.g. $\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) =
%\ParikhMap((abc)^*) = \CountOf{a} = \CountOf{b} = \CountOf{c} \land \CountOf{a}
%\geq 0$). This means they can be represented as a quantifier-free Presburger
%formula.
%
While Parikh's theorem applies to arbitrary context-free languages, in this
paper we focus only on regular languages.
